---
title: "Event demand and risk with monteriskr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Event demand and risk with monteriskr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.6)
library(monteriskr)
```

## The problem

An event planner wants to know not just the *expected* profit of a plan, but how
bad it could get. Two things make this harder than it looks:

* Attendance is a count, and it is **capped by venue capacity**. When an event
  sells out, the recorded attendance is only a lower bound on true demand.
* A plan is judged on risk (how likely is a loss? how large is the loss in a bad
  tail?), which needs a full distribution of outcomes, not a single forecast.

`monteriskr` fits a Bayesian negative binomial demand model that treats sold-out
events as right-censored, then simulates profit from the posterior and
summarises the risk.

## Data format

`fit_demand()` takes an ordinary data frame, for example one read with
`read.csv()`, with one row per past event:

| Column | Required | Notes |
|---|---|---|
| outcome (e.g. `attendance`) | yes | non-negative whole numbers, no blanks |
| covariates (e.g. `day_type`, `marketing`) | optional | no blanks; numeric ones not constant; factors need at least 2 levels |
| capacity | optional | one number, or the name of a column of positive values |

There are at least 8 rows (20 or more recommended). Rows where attendance
equals capacity are treated as sold out. Do not name a column `censored`: the
package uses that name internally.

The bundled `example_events` data has two extra columns (`true_demand`,
`sold_out`) that exist only because it is simulated; real data does not need
them.

```{r data}
head(example_events)
check_data(example_events, outcome = "attendance",
           covariates = c("day_type", "marketing"), capacity = 250)
```

## Fitting the demand model

Fitting compiles a Stan model, which takes about a minute the first time, so this
vignette loads a pre-computed fit. The call that produced it is shown in the
comment.

```{r fit}
# fit <- fit_demand(example_events, outcome = "attendance",
#                   covariates = c("day_type", "marketing"), capacity = 250,
#                   chains = 2, iter = 1000, seed = 1)
fit <- readRDS(system.file("extdata", "example_fit.rds", package = "monteriskr"))
round(summary(fit$brmsfit)$fixed[, 1:4], 3)
```

The coefficients are on the log scale: a weekend event has roughly
`exp(0.4)`, about 1.5 times, the expected demand of a weekday event, all else
equal. The default prior is `normal(0, 1)` on the coefficients.

## Simulating a plan

Describe the planned events as a data frame with the same covariate columns.
`simulate_risk()` draws a posterior parameter set for each run, simulates demand
for every event, caps it at capacity and derives revenue and profit. Prices and
costs are your assumptions, not model output.

```{r plan}
plan <- data.frame(
  day_type  = factor(c("weekday", "weekend", "weekend"),
                     levels = c("weekday", "weekend")),
  marketing = c(0, 1, 2)
)
risk <- simulate_risk(fit, plan, n_sims = 4000, price = 40,
                      variable_cost = 5, fixed_cost = 3000, seed = 2)
risk
```

```{r hist}
hist(risk$total_profit, breaks = 40, main = "Total profit across the 3 events",
     xlab = "Profit", col = "grey85", border = "white")
abline(v = mean(risk$total_profit), lwd = 2)
```

## Risk metrics

`risk_summary()` reports the headline numbers. Loss is `-profit`; VaR is the
`level`-quantile of loss and CVaR is the average loss at or beyond VaR (the
average of the worst `1 - level` share of outcomes). A negative VaR means a
profit even in the bad tail.

```{r summary}
risk_summary(risk, level = 0.95)
```

## Comparing scenarios

Run `simulate_risk()` once per alternative and compare. The first scenario is
the baseline.

```{r compare}
weekend <- data.frame(day_type = factor("weekend", levels = c("weekday", "weekend")),
                      marketing = 1)
cheap  <- simulate_risk(fit, weekend, n_sims = 4000, price = 30,
                        variable_cost = 5, fixed_cost = 3000, seed = 3)
pricey <- simulate_risk(fit, weekend, n_sims = 4000, price = 50,
                        variable_cost = 5, fixed_cost = 3000, seed = 3)
compare_scenarios(price_30 = cheap, price_50 = pricey)
```

Demand here does not respond to price, because price is not a model covariate:
price changes are assumptions you impose. If price varies in your history, add it
as a covariate and supply the planned price in `newdata`.

## Why censoring matters

Treating capped attendance as true demand makes the data look less variable than
it is and understates demand. A parameter-recovery study (50 simulated datasets
of 150 events per scenario, known true parameters; script in the package
source under `data-raw/`) compared the censored model with a naive one that
ignores capacity:

```{r validation}
val <- readRDS(system.file("extdata", "validation_results.rds", package = "monteriskr"))
s <- val$summary
s <- s[s$scenario == "heavy_censoring" & s$parameter != "Intercept",
       c("model", "parameter", "truth", "bias", "coverage90")]
s$bias <- round(s$bias, 3); s$coverage90 <- round(s$coverage90, 2)
knitr::kable(s, row.names = FALSE,
             caption = "About 25% of events sold out. Bias = mean(estimate - truth); coverage90 = share of 90% intervals containing the truth.")
```

With about a quarter of events sold out, the censored model is close to unbiased
while the naive model understates the weekend and marketing effects, overstates
the shape parameter, and its intervals rarely contain the truth. When almost
nothing sells out, the two agree. Fits that did not converge (Rhat above 1.05)
were excluded for both models and are counted in the stored results.

## Scope and limitations (v0.1)

* One count outcome, covariates, optional capacity censoring.
* No time-series structure; events are treated as exchangeable given covariates.
* Risk comes from posterior and demand uncertainty only; prices and costs are
  fixed inputs.
* `total_profit` sums profit over the events in `newdata`.
